Extended Abstract

### 2021-05-10 paper/paper.tex
Add variable typing rules
Write intro to inference

% Syntax types
\newcommand{\Fin}[1]{\type{Fin}~#1}
\newcommand{\sExpr}[1]{\type{Expr}~#1}
\newcommand{\sProc}[1]{\type{Proc}~#1}
\newcommand{\tvar}[2]{#1 ~\type{\ni_t}~ #2}
\newcommand{\tkind}[2]{#1 ~\type{\vdash_t}~ #2}
\newcommand{\ttype}[1]{\type{Type}~#1}
\newcommand{\tusage}[1]{\type{Usage}~#1}
\newcommand{\tCtx}[2]{\type{Ctx}_{#1}~#2}
\newcommand{\tSplit}[3]{#1~\type{=}~#2~\type{\uplus}~#3}
\newcommand{\tun}[1]{\type{un}~#1}
\newcommand{\tVar}[3]{#1 ~ \type{\ni} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tExpr}[3]{#1 ~ \type{\vdash} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tProc}[2]{#1 ~ \type{\vdash} ~ #2}

@@

% Type constructors
\newcommand{\tmvar}{\constr{mvar}}
\newcommand{\tchan}{\constr{chan}}
\newcommand{\tunit}{\constr{unit}}
\newcommand{\tsum}{\constr{sum}}
\newcommand{\tprod}{\constr{prod}}
\newcommand{\tzero}{\constr{0\cdot}}
\newcommand{\tone}{\constr{1\cdot}}
\newcommand{\tomega}{\constr{\omega\cdot}}

@@

A usage annotation $x$, type $t$ or context $\Gamma$ is unrestricted (shared, non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $x$ and $t$) \cite{Padovani15}.
\todo{typing judgment for variables}

We can finally define our typing judgment for expressions and processes. \todo{formation rules double lined}

\begin{mathpar}
  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ e : \sExpr{n} \\ t : \ttype{\gamma}}
             { \tExpr{\Gamma}{e}{t} }

@@

  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ p : \sProc{n}}
             { \tProc{\Gamma}{p} }

@@

  \inferrule { t : \ttype{\gamma} \\ \tProc{\Gamma , t}{p} } { \tProc{\Gamma}{\snew~p} }

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tchan~\tone~\tzero~t} \\ \tProc{\Theta,t}{p} }
             { \tProc{\Gamma}{\srecv~e~p} }

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tsum~s~t} \\ \tProc{\Theta,s}{p} \\ \tProc{\Theta,t}{q} }
             { \tProc{\Gamma}{\scase~e~p~q} }
\end{mathpar}

@@

Co-contextual type inference: traverses the syntax bottom up, collecting
constraints on the typing context.

Constraints can be of the following form:

\begin{itemize}
\item
  \(S = T\): types \(S\) and \(T\) must be unifiable.
\item
  \(S = T + R\): adding up the usage annotations in \(T\) and \(R\) must
  result in \(S\).
\end{itemize}

Constraints of the form \(S = T\) can always be eagerly unified. However
constraints of the form \(S = T + R\) might contain variables:
\(\texttt{mvar}~x = \texttt{mvar}~y + \texttt{mvar}~z\). In an open
process, these variables cannot safely be instantiated to \(0\). As
such, no substitution is possible, yet we need to remember the
constraint. (It might also be that some constraints have more than one
possible solution.) At the same time, we need to show that the
constraints that we postulate are enough to solve the problem.

>>>>

% Syntax types
\newcommand{\Fin}[1]{\type{Fin}~#1}
\newcommand{\fzero}{\constr{zero}}
\newcommand{\fsuc}{\constr{suc}}
\newcommand{\sExpr}[1]{\type{Expr}~#1}
\newcommand{\sProc}[1]{\type{Proc}~#1}
\newcommand{\tvar}[2]{#1 ~\type{\ni_t}~ #2}
\newcommand{\tkind}[2]{#1 ~\type{\vdash_t}~ #2}
\newcommand{\ttype}[1]{\type{Type}~#1}
\newcommand{\tusage}[1]{\type{Usage}~#1}
\newcommand{\tCtx}[2]{\type{Ctx}_{#1}~#2}
\newcommand{\tSplit}[3]{#1~\type{=}~#2~\type{\uplus}~#3}
\newcommand{\tun}[1]{\type{un}~#1}
\newcommand{\tVar}[3]{#1 ~ \type{\ni} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tExpr}[3]{#1 ~ \type{\vdash} ~ #2 ~ \type{:} ~ #3}
\newcommand{\tProc}[2]{#1 ~ \type{\vdash} ~ #2}

@@

% Typing judgment constructors
\newcommand{\tmvar}{\constr{mvar}}
\newcommand{\tchan}{\constr{chan}}
\newcommand{\tunit}{\constr{unit}}
\newcommand{\tsum}{\constr{sum}}
\newcommand{\tprod}{\constr{prod}}
\newcommand{\tzero}{\constr{0\cdot}}
\newcommand{\tone}{\constr{1\cdot}}
\newcommand{\tomega}{\constr{\omega\cdot}}

% Constraint constructors
\newcommand{\eqconstr}[2]{#1 ~ \constr{\stackrel{c}{=}} ~ #2}
\newcommand{\sumconstr}[3]{#1 ~ \constr{\stackrel{c}{=}} ~ #2 ~ \constr{+} ~ #3}

@@

A usage annotation $x$, type $t$ or context $\Gamma$ is unrestricted (shared, non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $x$ and $t$) \cite{Padovani15}.

We can finally define our typing judgment for variables, expressions and processes -- some are omitted for brevity.

\begin{mathpar}
%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ i : \Fin{n} \\ t : \ttype{\gamma} }
%             { \tVar{\Gamma}{i}{t} }

  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ \tun{\Gamma} \\ t : \ttype{\gamma} }
             { \tVar{\Gamma,t}{\fzero}{t} }
             \rulename{zero}

  \inferrule { \tVar{\Gamma}{i}{t} \\ s : \ttype{\gamma} \\ \tun{s} }
             { \tVar{\Gamma,s}{\fsuc~i}{t} }
             \rulename{suc}

%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ e : \sExpr{n} \\ t : \ttype{\gamma}}
%             { \tExpr{\Gamma}{e}{t} }

@@

%  \inferrule { \Gamma : \tCtx{n}{\gamma} \\ p : \sProc{n}}
%             { \tProc{\Gamma}{p} }

@@

  \inferrule { t : \ttype{\gamma} \\ \tProc{\Gamma , t}{p} } { \tProc{\Gamma}{\snew~p} }
             \rulename{new}

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tchan~\tone~\tzero~t} \\ \tProc{\Theta,t}{p} }
             { \tProc{\Gamma}{\srecv~e~p} }
             \rulename{recv}

  \inferrule { \tSplit{\Gamma}{\Delta}{\Theta} \\ \tExpr{\Delta}{e}{\tsum~s~t} \\ \tProc{\Theta,s}{p} \\ \tProc{\Theta,t}{q} }
             { \tProc{\Gamma}{\scase~e~p~q} }
             \rulename{case}
\end{mathpar}

@@

Co-contextual type inference traverses processes bottom up collecting constraints.
Limiting the information flow to being bottom up has the advantage of making inference subproblems independent and thus easily parallelisable.
For the shared \picalc{}, inference constraints amount to unification problems that can always be eagerly solved.
The linear \picalc{} however requires that some constraints are deferred to a later stage, when more information is available.

Let us suppose we can safely instantiate to $\tzero$ every usage annotation that is unrestricted.
Consider the open process $\ssend~a~\sunit~(\ssend~x~a~\send)$ where $x$ and $a$ are free: we send the unit over $a$, then send $a$ over $x$ and terminate.
By the time we reach $\send$ both $x$ and $a$ must be unrestricted, and thus channel $x$ must be of some type $\tchan~\tzero~\tone~t$.
Since we send a unit over $a$, $a$ must be of some type $s$ such that $\tSplit{s}{\tchan~\tzero~\tone~\tunit}{t}$.
From this we can infer that both $s$ and $t$ must be a channel type and thus $\tSplit{s_i}{\tzero}{t_i}$ and $\tSplit{s_o}{\tone}{t_o}$ for some $s_i$, $t_i$, $s_o$ and $t_o$, but we cannot infer their usage annotations: all we can deduce is their lower bounds $\tzero$ and $\tone$.
We must defer the resolution of these constraints until more information becomes available about the context of this open process.

We present two kinds of constraints, showcase how to encode inference problems for some of the typing rules in \autoref{type-system}, and state the lemmas that will allow us to show that our approach to inference is both sound and complete.

Constraints take two forms: the binary $\eqconstr{S}{T}$, where $S$ and $T$ are meant to be unified, and the ternary $\sumconstr{S}{T}{R}$, where the usage annotations in $T$ and $R$ are meant to be added up to $S$.
Both these constraints are defined on usage annotations, types, and typing contexts.
Constraints of the former form can \emph{always} be eagerly solved, while constraints of the latter \emph{sometimes} cannot, and must be deferred.
